*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-23 Code/Manuscript_O2.Rmd
more edits

---
title: "*Prochlorococcus marinus* responses to light and oxygen" 
author:
  - name: '**Mireille Savoie**'
    affil: 1
    main: true
    email: msavoie@mta.ca
  - name: Aurora Mattison
    affil: 1, 2
  - name: Julie Nadeau
    affil: 1
  - name: Laurel Genge
    affil: 1
  - name: Sylwia Śliwińska-Wilczewska
    affil: 1
  - name: Maximilian Berthold
    affil: 1, 3
  - name: Naaman Omar
    affil: 1
  - name: Ondřej Prášil
    affil: 1, 4
    main: false
  - name: Amanda M. Cockshutt
    affil: 1, 5
    main: false
  - name: Douglas A. Campbell
    affil: 1
    main: false
    orcid: '0000-0001-8996-5463'
    email: 'dcampbel@mta.ca'
affiliation:
  - num: 1
    address: Department of Biology, Mount Allison University
institute:
  - mta: 'Department of Biology, Mount Allison University, 63B York St., Sackville NB, Canada, E4L 1G7.'
  - num: 2
  - ubc: 'Department of Biology, University of British Columbia'
  - num: 3
  - umwelt: 'Umweltbundesamtm, Dessau-Roßlau, Germany'
  - num: 4
  - alga: 'Algatech, Czech Academy of Science'
  - num: 5
  - sfx: 'Department of Chemistry, St. Frances Xavier University'
date: "`r format(Sys.Date())`"
output:
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  bookdown::word_document2:
    code_folding: show
    reference_docx: KnitTemplate.docx
    keep_md: yes
    toc: FALSE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
always_allow_html: true
bibliography: [Manuscript_O2.bib, RPackages.bib]
csl: plos-one.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

where ΔOD~max~ is maximum ΔOD, ΔOD~min~ is minimum ΔOD, t is time
duration over the growth trajectory.  Annotated code for data import, transformations and analyses is available at XXXXX 
CITAtIONS of github locations of code etc. XXXXX

@@

The applied photoperiods were delivered using the sine circadian light function of the PSI Multicultivator to simulate light exposure approximating sun rise through to sunset. The area under the sinusoidal curves is equivalent to the area of a triangular photoregime of equivalent photoperiod (Campbell, unpub), therefore the equation to determine the cumulative diel PUR (µmol photons m^-2^ d^-1^) is one half of the base (photoperiod) multiplied by the height (PUR) (equation \@ref(eq:SumPURDay)),

@@

Under 250 µM O~2~ the growth rates of *P. marinus* MED4, Clade HLI, increased with higher imposed PAR and longer photoperiods
(Figure \@ref(fig:MED4PAR)), across all spectral wavebands. No growth was observed under any imposed conditions under a 4 h photoperiod. The maximum growth rate (µ~max~) was `r PURPARFits %>% filter(Strain == "MED4", WL == "450", O2_uM == "250", Photoperiod == 16) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^ achieved under 180 uE blue light and 16 h photoperiod.

Similar to growth trends under 250 µM O~2~, MED4 maintained at 25 µM O~2~ showed fastest growth when the
photoperiod was 16 h for each spectral waveband, across PAR levels (Figure \@ref(fig:MED4PAR)).
The µ~max~ was `r PURPARFits %>% filter(Strain == "MED4", WL == "450", O2_uM == "25", Photoperiod == 16) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^  achieved under 180 µmol photons m^-2^ s^-1^ blue light and 16 h photoperiod. The 4 h photoperiod
experiments under white LED light were not performed because no growth was achieved when grown under an 8 h photoperiod of white LED light.

@@

The GAM model in Figure \@ref(fig:MED4GAM) summarizes MED4 growth responses to red (A,B) or blue (C,D) peak PAR and photoperiod across 2 imposed oxygen concentrations.  Under 250 µM O~2~ MED4 achieved fastest growth rates above peak blue light of ~180 µmol photons m^-2^ s^-1^, and the longest photoperiod of 16, indicated by the 0.64 d^-1^ contour line representing the 90^th^ percentile of maximum achieved growth rate (Figure \@ref(fig:MED4GAM)C).  Growth decreased with decreasing photoperiod and decreasing peak PAR. Under red light growth was generally slower but the pattern of growth responses to photoperiod and PAR was similar (Figure \@ref(fig:MED4GAM)A).  Note the exclusion of MED4 from growth under 4 h photoperiod under both red and blue light \@ref(fig:MED4GAM). Under 25 µM O~2~ MED4 showed similar growth responses, but was excluded from both 4 and 8 h photoperiods.  MED4 did not grow under 25 µM O~2~, so no GAM model was run. 
Considering the range of PAR levels, and spectral bands that MED4 can utilize, MED4 can inhabit not just shallow depths, where light levels are high, but also deeper regions, characterized by lower level of blue light, subject to the limitation of a photoperiod of more than 4 h, even after depth attenuation of light. The photoregimes of winter temperate zones, due to shorter photoperiods, exclude MED4 from growth at any depth, however temperate photoperiods and light levels for the remainder of the year are potentially adequate to support MED4 growth, if water temperatures warm into the Clade HLI tolerance range.

@@

Under 250 µM O~2~ growth rates of *P. marinus* SS120, Clade LLIII, increased with longer photoperiods, under 30 µmol photons m^-2^ s^-1^ peak PAR and (Figure \@ref(fig:SS120PAR)). No growth was observed under any blue light photoperiods when exposed to peak PAR of 90 µmol photons m^-2^ s^-1^ or greater. Growth rate, however increased with increasing photoperiods for white and red light under peak PAR of 90 µmol photons m^-2^ s^-1^.  Growth rate decreased with longer photoperiods and showed growth inhibition at 12 and 16 h photoperiods under PAR of 180 µmol photons m^-2^ s^-1^  white, red or blue LED light.  The µ~max~ was `r PURPARFits %>% filter(Strain == "SS120", WL == "WW", O2_uM == "250", Photoperiod == 16) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^  achieved under 90 µmol photons m^-2^ s^-1^ white LED light and 16 h photoperiod.

@@

The GAM model in Figure \@ref(fig:SS120GAM) summarizes growth responses of SS120 to red (A) or blue (B) peak PAR and photoperiod, across the 2 imposed oxygen concentrations. Under 250 µM O~2~, Figure \@ref(fig:SS120GAM)A showed highest growth rates below PAR of 50 µmol photons m^-2^ s^-1^ and photoperiods between 8 and 12 h, indicated by the contour line labeled 0.29 d^-1^ (representing the 90^th^ percentile of achieved growth rate).  Under 250 µM O~2~ SS120 is constrained to deeper ocean waters through its intolerance of higher blue PAR levels. These findings align with @mooreComparativePhysiologySynechococcus1995 and are expected for a low light clade.  The disjunct regions of the GAM plot results from variable growth success of SS120 under 250 µM O~2~.  Growth rate patterns under red light were similar, although somewhat faster. In contrast, under 25 µM O~2~ and a photoperiod of 8 h SS120 exploited all blue peak PAR levels, achieving faster growth rates at a higher PAR of \~100 µmol photons m^-2^ s^-1^, indicated by the contour line labeled 0.5 d^-1^ (representing the 90^th^ percentile of achieved growth rate), out pacing the 90^th^ percentile fastest growth rates under 250 µM O~2~. Under red light and 25 µM O~2~ light SS120 grew across most conditions of peak PAR and photoperiod, achieving fastest growth under long photoperiods and peak PAR below 50 \~100 µmol photons m^-2^ s^-1^.  Thus, the designation of SS120 as a LL strain is dependent upon the [O~2]. SS120 did not, however, grow reliably under tested conditions at 2.5 µM O~2~.

@@

For MIT9313 under 25 µM O~2~ growth rate increased with increasing photoperiods for all spectral wavebands tested (Figure \@ref(fig:MIT9313PAR)), with the fastest overall growth rate for MIT9313 
`r PURPARFits %>% filter(Strain == "MIT9313", WL == "450", O2_uM == "25", Photoperiod == 16, Par_ue == 90) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^ achieved under peak PAR of 90 µmol photons m^-2^ s^-1^ and 16 h blue light photoperiod. In marked contrast to the 250 µM O~2~ growth experiments, MIT9313 grew when exposed to peak PAR of 180 µmol photons m^-2^ s^-1^ and blue light under all photoperiods except 16 h; additionally, white LED and red light treatments induced growth across all tested photoperiods under 25 µM O~2~. The 25 µM O~2~, 4 h photoperiod experiments under white LED light and were not performed due to time constraints.

MIT9313 grew under 2.5 µM O~2~  particularly under blue LED light, albeit generally slower than under the parallel experiments at 25 µM O~2~  (Figure \@ref(fig:MIT9313PAR)). Growth estimates showed scatter among replicates, suggesting  2.5 µM O~2~  is near the tolerance limit for growth of MIT9313. Growth rates increased with longer photoperiods under blue light treatments under peak PAR of 90 µmol photons m^-2^ s^-1^ . Growth  for MIT9313 under PAR of 180 µmol photons m^-2^ s^-1^ and blue light treatment decreased with increasing photoperiods with full growth inhibition under a 16 h photoperiod. The red light peak PAR of 180 µmol photons m^-2^ s^-1^ showed similar growth rates for 8 and 12 h photoperiods. The µ~max~ was `r PURPARFits %>% filter(Strain == "MIT9313", WL == "450", O2_uM == "2.5", Photoperiod == 4, Par_ue == 180) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^ achieved under 4 h blue light photoperiod and PAR of 180 µmol photons m^-2^ s^-1^.  XXXXTHIS DOES NOT LOOK correct; 12 h, 90 uE photons blue light looks higher. The 2.5 µM O~2~ white LED treatments under 4, 8 and 16 h photoperiods and red light under 4 and 16 h photoperiods were not performed due to time constraints.

@@

The GAM model in Figure \@ref(fig:MIT9313GAM) summarizes MIT9313 growth responses to red (A) or blue (B) peak PAR and photoperiod. Under 250 µM O~2~, Figure \@ref(fig:MIT9313GAM)A shows MIT9313 achieves fastest growth rates between blue peak PAR of 30 µmol photons m^-2^ s^-1^ and 50 µmol photons m^-2^ s^-1^ and photoperiods longer than 8 h, indicated by the contour line labeled 0.51 d^-1^ representing the 90^th^ percentile of achieved growth rates. Figure \@ref(fig:MIT9313GAM)A also shows that growth rate increases with longer photoperiods, as long as the blue peak PAR levels remain below 50 µmol photons m^-2^ s^-1^.  In contrast, under red light and 250 µM O~2~ MIT9313 grows faster while exploiting higher peak PAR and longer photoperiods.  Figure \@ref(fig:MIT9313GAM)C shows that MIT9313 maintains growth even under 2.5 µM O~2~, under photoperiods between 4 and 8 h and peak blue PAR between 50 to 10 µmol photons m^-2^ s^-1^ PAR (Figure \@ref(fig:MIT9313GAM)C).  Thus the designation of MIT9313 as a LL clade is dependent upon [O2] and light spectra. (Figure \@ref(fig:MIT9313GAM)B).

@@

# Summary and Conclusions {.unnumbered}

>>>>

---
title: "*Prochlorococcus marinus* responses to light and oxygen" 
author:
  - name: '**Mireille Savoie**'
    institute: mta
    main: true
    email: msavoie@mta.ca
  - name: Aurora Mattison
    institute: [mta, ubc]
  - name: Julie Nadeau
    institute: mta
  - name: Laurel Genge
    institute: mta
  - name: Sylwia Śliwińska-Wilczewska
    institute: [mta, other]
  - name: Maximilian Berthold
    institute: mta
  - name: Naaman Omar
    institute: mta
  - name: Ondřej Prášil
    institute: [mta, algatech]
    main: false
  - name: Amanda M. Cockshutt
    institute: [mta, sfx]
    main: false
  - name: Douglas A. Campbell
    institute: mta
    affil: 1
    main: false
    orcid: '0000-0001-8996-5463'
    email: 'dcampbel@mta.ca'
affiliation:
  - num: 1
    address: Department of Biology, Mount Allison University
institute:
  - mta: 'Department of Biology, Mount Allison University, 63B York St., Sackville NB, Canada, E4L 1G7.'
  - ubc: 'Department of Biology, University of British Columbia'
  - sfx: 'Department of Chemistry, St. Frances Xavier University'
  - algatech: 'Institute of Microbiology, Center Algatech, Laboratory of Photosynthesis, Novohradska 237, Trebon, CZ 37981, Czech Republic.'

date: "`r format(Sys.Date())`"
output:
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  bookdown::word_document2:
    code_folding: show
    reference_docx: KnitTemplate.docx
    keep_md: yes
    toc: FALSE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
always_allow_html: true
bibliography: [Manuscript_O2.bib, RPackages.bib]
csl: plos-one.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

where ΔOD~max~ is maximum ΔOD, ΔOD~min~ is minimum ΔOD, t is time
duration over the growth trajectory.

Figure \@ref(fig:deltaODLogGrowthPlot) is an example of chlorophyll specific
growth estimates fitted from the high resolution ΔOD measurements for
each tube in a Multicultivator. The residuals of the logistic growth
curve fit are shown and the growth spectral waveband is plotted and
illustrates the imposed PAR (µmol photons m^-2^ s^-1^) and photoperiod
(h). Annotated code for data import, transformations and analyses is available at https://github.com/FundyPhytoPhys/prochlorococcus_o2/tree/main/Data/RawData

@@

The applied photoperiods were delivered using the sine circadian light function of the PSI Multicultivator to simulate light exposure approximating sun rise through to sunset. The area under the sinusoidal curves is equivalent to the area of a triangular photoregime of equivalent photoperiod (Campbell, unpub), therefore the equation to determine the cumulative diel PUR (µmol photons m^-2^ d^-1^) is one half of the base (photoperiod) multiplied by the height (PUR) (Equation \@ref(eq:SumPURDay)),

@@

Under 250 µM O~2~ the growth rates of *P. marinus* MED4, Clade HLI, increased with higher imposed PAR and longer photoperiods
(Figure \@ref(fig:MED4PAR)), across all spectral wavebands. No growth was observed under any imposed conditions under a 4 h photoperiod. The maximum growth rate (µ~max~) was `r PURPARFits %>% filter(Strain == "MED4", WL == "450", O2_uM == "250", Photoperiod == 16) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^ achieved under 180 µE blue light and 16 h photoperiod.

Similar to growth trends under 250 µM O~2~, MED4 maintained at 25 µM O~2~ showed fastest growth when the
photoperiod was 16 h for each spectral waveband, across PAR levels (Figure \@ref(fig:MED4PAR)).
The µ~max~ was `r PURPARFits %>% filter(Strain == "MED4", WL == "450", O2_uM == "25", Photoperiod == 16) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^ (Table. \@ref(tab:MaxGrowthTable)) achieved under 180 µmol photons m^-2^ s^-1^ blue light and 16 h photoperiod. The 4 h photoperiod
experiments under white LED light were not performed because no growth was achieved when grown under an 8 h photoperiod of white LED light.

@@

The GAM model in Figure \@ref(fig:MED4GAM) summarizes MED4 growth responses to red (A,B) or blue (C,D) peak PAR and photoperiod across 2 imposed oxygen concentrations.  Under 250 µM O~2~ MED4 achieved fastest growth rates above peak blue light of ~180 µmol photons m^-2^ s^-1^, and the longest photoperiod of 16, indicated by the 0.64 d^-1^ contour line representing the 90^th^ percentile of maximum achieved growth rate (Figure \@ref(fig:MED4GAM)C).  Growth decreased with decreasing photoperiod and decreasing peak PAR. Under red light growth was generally slower but the pattern of growth responses to photoperiod and PAR was similar (Figure \@ref(fig:MED4GAM)A).  Note the exclusion of MED4 from growth under 4 h photoperiod under both red and blue light (Figure \@ref(fig:MED4GAM)). Under 25 µM O~2~ MED4 showed similar growth responses, but was excluded from both 4 and 8 h photoperiods.  MED4 did not grow under 2.5 µM O~2~, so no GAM model was run. 
Considering the range of PAR levels, and spectral bands that MED4 can utilize, MED4 can inhabit not just shallow depths, where light levels are high, but also deeper regions, characterized by lower level of blue light, subject to the limitation of a photoperiod of more than 4 h, even after depth attenuation of light. The photoregimes of winter temperate zones, due to shorter photoperiods, exclude MED4 from growth at any depth, however temperate photoperiods and light levels for the remainder of the year are potentially adequate to support MED4 growth, if water temperatures warm into the Clade HLI tolerance range.

@@

Under 250 µM O~2~ growth rates of *P. marinus* SS120, Clade LLIII, increased with longer photoperiods, under 30 µmol photons m^-2^ s^-1^ peak PAR and (Figure \@ref(fig:SS120PAR)). No growth was observed under any blue light photoperiods when exposed to peak PAR of 90 µmol photons m^-2^ s^-1^ or greater. Growth rate, however increased with increasing photoperiods for white and red light under peak PAR of 90 µmol photons m^-2^ s^-1^.  Growth rate decreased with longer photoperiods and showed growth inhibition at 12 and 16 h photoperiods under PAR of 180 µmol photons m^-2^ s^-1^  white, red or blue LED light.  The µ~max~ was `r PURPARFits %>% filter(Strain == "SS120", WL == "WW", O2_uM == "250", Photoperiod == 16) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^ (Table. \@ref(tab:MaxGrowthTable)) achieved under 90 µmol photons m^-2^ s^-1^ white LED light and 16 h photoperiod.

@@

The GAM model in Figure \@ref(fig:SS120GAM) summarizes growth responses of SS120 to red (A,B) or blue (C,D) peak PAR and photoperiod, across the 2 imposed oxygen concentrations. Under 250 µM O~2~, Figure \@ref(fig:SS120GAM)C showed highest growth rates below PAR of 50 µmol photons m^-2^ s^-1^ and photoperiods between 8 and 12 h, indicated by the contour line labeled 0.19 d^-1^ (representing the 90^th^ percentile of achieved growth rate).  Under 250 µM O~2~ SS120 is constrained to deeper ocean waters through its intolerance of higher blue PAR levels. These findings align with @mooreComparativePhysiologySynechococcus1995 and are expected for a low light clade.  The disjunct regions of the GAM plot results from variable growth success of SS120 under 250 µM O~2~.  Growth rate patterns under red light were similar, although somewhat faster. In contrast, under 25 µM O~2~ and a photoperiod of 8 h SS120 exploited all blue peak PAR levels, achieving faster growth rates at a higher PAR of \~100 µmol photons m^-2^ s^-1^, indicated by the contour line labeled 0.4 d^-1^ (representing the 90^th^ percentile of achieved growth rate), out pacing the 90^th^ percentile fastest growth rates under 250 µM O~2~. Under red light and 25 µM O~2~ (Figure \@ref(fig:SS120GAM)B) SS120 grew across most conditions of peak PAR and photoperiod, achieving fastest growth under long photoperiods and peak PAR below 50 \~100 µmol photons m^-2^ s^-1^.  Thus, the designation of SS120 as a LL strain is dependent upon the [O~2~]. SS120 did not, however, grow reliably under tested conditions at 2.5 µM O~2~.

@@

For MIT9313 under 25 µM O~2~ growth rate increased with increasing photoperiods for all spectral wavebands tested (Figure \@ref(fig:MIT9313PAR)), with the fastest overall growth rate for MIT9313 
`r PURPARFits %>% filter(Strain == "MIT9313", WL == "WW", O2_uM == "25", Photoperiod == 16, Par_ue == 90) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^ achieved under peak PAR of 90 µmol photons m^-2^ s^-1^ and 16 h white LED light photoperiod. In marked contrast to the 250 µM O~2~ growth experiments, MIT9313 grew when exposed to peak PAR of 180 µmol photons m^-2^ s^-1^ and blue light under all photoperiods except 16 h; additionally, white LED and red light treatments induced growth across all tested photoperiods under 25 µM O~2~. The 25 µM O~2~, 4 h photoperiod experiments under white LED light and were not performed due to time constraints.

MIT9313 grew under 2.5 µM O~2~  particularly under blue LED light, albeit generally slower than under the parallel experiments at 25 µM O~2~  (Figure \@ref(fig:MIT9313PAR)). Growth estimates showed scatter among replicates, suggesting  2.5 µM O~2~  is near the tolerance limit for growth of MIT9313. Growth rates increased with longer photoperiods under blue light treatments under peak PAR of 90 µmol photons m^-2^ s^-1^ . Growth  for MIT9313 under PAR of 180 µmol photons m^-2^ s^-1^ and blue light treatment decreased with increasing photoperiods with full growth inhibition under a 16 h photoperiod. The red light peak PAR of 180 µmol photons m^-2^ s^-1^ showed similar growth rates for 8 and 12 h photoperiods. The µ~max~ was `r PURPARFits %>% filter(Strain == "MIT9313", WL == "450", O2_uM == "2.5", Photoperiod == 12, Par_ue == 90) %>% select(ReplicateMean_deltaOD_Lmu_corr_day) %>% max() %>% round(digits = 2)` d^-1^ achieved under 12 h blue light photoperiod and PAR of 90 µmol photons m^-2^ s^-1^. The 2.5 µM O~2~ white LED treatments under 4, 8 and 16 h photoperiods and red light under 4 and 16 h photoperiods were not performed due to time constraints.

@@

The GAM model in Figure \@ref(fig:MIT9313GAM) summarizes MIT9313 growth responses to red (A,B,C) or blue (D,E,F) peak PAR and photoperiod. Under 250 µM O~2~, Figure \@ref(fig:MIT9313GAM)D shows MIT9313 achieves fastest growth rates between blue peak PAR of 30 µmol photons m^-2^ s^-1^ and 50 µmol photons m^-2^ s^-1^ and photoperiods longer than 8 h, indicated by the contour line labeled 0.52 d^-1^ representing the 90^th^ percentile of achieved growth rates. Figure \@ref(fig:MIT9313GAM)D also shows that growth rate increases with longer photoperiods, as long as the blue peak PAR levels remain below 50 µmol photons m^-2^ s^-1^.  In contrast, under red light and 250 µM O~2~ MIT9313 grows faster while exploiting higher peak PAR and longer photoperiods.  Figure \@ref(fig:MIT9313GAM)F shows that MIT9313 maintains growth even under 2.5 µM O~2~, under photoperiods between 4 and 8 h and peak blue PAR between 50 to 10 µmol photons m^-2^ s^-1^ PAR.  Thus the designation of MIT9313 as a LL clade is dependent upon [O~2~] and light spectra. (Figure \@ref(fig:MIT9313GAM)E).

@@

# Summary and conclusions {.unnumbered}
